Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Edits to start of methodology and GLQ
Minor changes to the text.

We present a new algorithm for computing the gravitational fields generated by a
tesseroid with an arbitrary continuous density function on an external point.
It is based on the three dimensional GLQ integration, a two dimensional version of the
adaptive discretization of \citet{Uieda2016} (following \citet{Lin2018}),
and a new density-based radial discretization algorithm.
To ensure the accuracy of the numerical approximation, we empirically determine
optimal values for the controlling parameters by comparing the numerical results with
analytical solutions for spherical shells.
Finally, we apply the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with linear and exponentially increasing density with depth.

@@

Consider an arbitrary tesseroid in a geocentric spherical coordinate system defined by
pairs of geocentric latitudinal ($\phi_1$, $\phi_2$), longitudinal ($\lambda_1$,
$\lambda_2$) and radial ($r_1$, $r_2$) boundaries.
We define an external computation point $P(r, \phi, \lambda)$ located at radius $r$,
geocentric latitude $\phi$, and longitude $\lambda$ where the gravitational fields the
tesseroid generates are going to be calculated.
The first- and second-order derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system of $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first- and second-order derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary continuous function $\rho(r)$.
Thus, the integrals for the gravitational fields are slightly modified to

@@

\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}

\noindent in which $\alpha, \beta \in \{x, y, z\}$, $\delta_{\alpha\beta}$ is
Kronecker's delta, $G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the
gravitational constant and

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') f(r', \phi', \lambda')
    dr' d\phi' d\lambda' \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

\begin{equation}
    A =
    \frac{(\lambda_2 - \lambda_1)(\phi_2 - \phi_1)(r_2 - r_1)}{8},
\end{equation}

\noindent $f(r', \phi', \lambda')$ is the non-density dependent portion of the
integration kernels, $(r_i, \phi_j, \lambda_k)$ are the quadrature nodes, and
$W_i^r$, $W_j^\phi$, $W_k^\lambda$ are the quadrature weights.
The GLQ nodes and weights can be calculated as \citet[p.~391]{Hildebrand1987} (see
\citet{Uieda2016} for nodes and weights definitions on similar notation used on previous
equations).

It's worth noting that a GLQ is equivalent to approximate the tesseroid by $N^r \times
N^\phi \times N^\lambda$ point masses located on the quadrature nodes
\citep{Ku1977, Li2011, Uieda2016}.

@@

\citet{Fukushima2018} propose a generalization of his method to polynomial density
functions.
On the other hand, our method is not bounded to a specific density function.
Nevertheless, the density-based discretization algorithm could be used before applying
the numerical integration developed by \citet{Fukushima2018} assigning a low polynomial
order density to each discretizing tesseroid.
Although this method would involve a computational error due to the approximation of the
density function as a polynomial one, it's feasibly if we account for this error and
a method to limit it.

>>>>

We present a new algorithm for computing the gravitational potential and its gradient
generated by a tesseroid with an arbitrary continuous density function on an external
point.
It is based on the three dimensional GLQ integration, a two dimensional version of the
adaptive discretization of \citet{Uieda2016} (following \citet{Lin2018}),
and a new density-based radial discretization algorithm.
To ensure the accuracy of the numerical approximation, we empirically determine
optimal values for the controlling parameters by comparing the numerical results with
analytical solutions for spherical shells.
Finally, we apply the methodology to model the Neuqu\'en basin, Argentina, using
tesseroids with linear and exponentially increasing density with depth.

@@

Consider a tesseroid in a geocentric spherical coordinate system defined by
pairs of geocentric latitudinal ($\phi_1$, $\phi_2$), longitudinal ($\lambda_1$,
$\lambda_2$), and radial ($r_1$, $r_2$) boundaries.
We define an external computation point $P(r, \phi, \lambda)$ located at radius $r$,
geocentric latitude $\phi$, and longitude $\lambda$.
The derivatives of the gravitational potential are taken with
respect to the local north-oriented Cartesian coordinate system with origin at $P$
(Fig.~\ref{fig:tesseroid}).
\citet{Grombein2013} provide efficient formulations for the volume integrals of the
gravitational potential and its first- and second-order derivatives of a tesseroid with
homogeneous density.
Here, we will assume that the tesseroid has a density varying with $r$ according to an
arbitrary continuous function $\rho(r)$.
Thus, the integrals for the gravitational potential and its first-order derivatives are
slightly modified to

@@

\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\\
\noindent in which $\alpha \in \{x, y, z\}$,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant
and

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \rho(r') f(r', \phi', \lambda')
        dr' d\phi' d\lambda' \approx& \\
        A
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        W_i^r W_j^\phi W_k^\lambda
        &\rho(r_i) f(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') f(r', \phi', \lambda')
    dr' d\phi' d\lambda' \approx
    A
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) f(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}
\\
\noindent where

\begin{equation}
    A = \frac{(\lambda_2 - \lambda_1)(\phi_2 - \phi_1)(r_2 - r_1)}{8},
\end{equation}
\\
\noindent $f(r', \phi', \lambda')$ is an integral kernel for a homogeneous tesseroid
\citep{Grombein2013},
$(r_i, \phi_j, \lambda_k)$ are the coordinates of the quadrature nodes,
$N^r$, $N^\phi$, $M^\lambda$ are the quadrature orders and $W_i^r$, $W_j^\phi$,
$W_k^\lambda$ are the quadrature weights in the radial, latitudinal, and longitudinal
directions, respectively.
\citet[p.~391]{Hildebrand1987} provides the formulas for calculating the GLQ weights and
pre-computed values of the Legendre polynomial roots for low orders.
See \citet{Uieda2016} for a more detailed description using a similar notation to the
one used here.
It is worth noting that a GLQ is equivalent to approximating the tesseroid by $N^r
\times N^\phi \times N^\lambda$ point masses located on the quadrature nodes
\citep{Ku1977, Asgharzadeh2007}.

@@

\citet{Fukushima2018} propose a generalization of his method to polynomial density
functions.
On the other hand, our method is not bounded to a specific density function.
Nevertheless, the density-based discretization algorithm could be used before applying
the numerical integration developed by \citet{Fukushima2018} assigning a low polynomial
order density to each discretizing tesseroid.
Although this method would involve a computational error due to the approximation of the
density function as a polynomial one, its feasibly if we account for this error and
a method to limit it.
